---
title: "Processing with 16S Hawaii"
author: "Erik Killian"
date: "2025-08-05"
output: html_document
---
---
title: "phyloseq_processing"
author: "Erik Killian"
date: "2025-06-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

This follows the same protocol as max_EE_2, except now I've fixed the Hawaii data that was previously a copy of the ITS data instead of 16S data.


# Step 1: Load Phyloseq Pieces
```{r LOAD HERE}
library(dada2)
library(phyloseq)
library(readxl)
library(dplyr)
library(ggplot2)

getwd()
# Loads merged field data - created in the chunk below.
fieldData = readRDS("C:/Users/killi/OneDrive - Montana State University/PostDoc/Microbiome_bacteria/fieldData_Combined.RDS")

# Loaded from tempest output 
## 2021.R and 2022.R create seqtabs 
## merging.R merge seqtabs, remove chimeras, and creates taxtab
## Phylogenetic_Tree.R uses merging.R output "seqtab_NoC" to create phylo tree

## seqtab is the merged seqtabs from each location-year, merged then chimeras removed.
## taxtab is the taxonomy table, made from seqtab to classify taxonomy.

seqtab = readRDS("Phyloseq2.0_Tempest/16S_maxEE_2_HI16S/seqtab.merged.NoC.RDS")
taxtab = readRDS("Phyloseq2.0_Tempest/16S_maxEE_2_HI16S/taxTab.RDS")

```

# Step 2: Merge Phyloseq into One

```{r Create phyloseq object}
ps = phyloseq(otu_table(seqtab,taxa_are_rows = FALSE),
              tax_table(taxtab),
              sample_data(fieldData))


# Add ASV short names
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps.all = ps
ps = subset_taxa(ps, Kingdom == "Bacteria") # Removes only 403 taxa 

saveRDS(ps,"Phyloseq2.0_Tempest/16S_maxEE_2_HI16S/ps.RDS")

table(taxtab[,"Kingdom"])
table(tax_table(ps)[,"Kingdom"])
table(sample_data(ps)[,"Location"])

```

# Step 3: Add Environmental Data
```{r Add Env data}
# Add soil texture, precip, previous crop, OM
#readRDS("ps.RDS")
psField = ps
# Load environmental datasets
bz.soil =readxl::read_xlsx("soilsummary.xlsx",sheet = 3)
bz.soil = bz.soil[!is.na(bz.soil$`Kind Of Sample`),]
colnames(bz.soil)[colnames(bz.soil) == "Sample ID"] = "Block"

bz22.soil = readxl::read_xlsx("soilsummary.xlsx", sheet = 4)
bz22.soil = bz22.soil[!is.na(bz22.soil$`Kind Of Sample`),]
colnames(bz22.soil)[colnames(bz22.soil) == "Sample ID"] = "Block"

mo21.soil = readxl::read_xlsx("soilsummary.xlsx", sheet = 5)
mo21.soil = mo21.soil[!is.na(mo21.soil$City),]
colnames(mo21.soil)[colnames(mo21.soil) == "Sample ID"] = "Block"
mo21.soil$Block = 1:12

mo22.soil = readxl::read_xlsx("soilsummary.xlsx", sheet = 6)
mo22.soil = mo22.soil[!is.na(mo22.soil$`Kind Of Sample`),]
colnames(mo22.soil)[colnames(mo22.soil) == "Sample ID"] = "Block"
mo22.soil$Block = 1:12

sd21.soil = readxl::read_xlsx("soilsummary.xlsx", sheet = 7)
sd22.soil = readxl::read_xlsx("soilsummary.xlsx", sheet = 8)


x = sample_data(psField)
x = as.data.frame(as.matrix(x))
x <- x %>%
  mutate(Block = ifelse(grepl("B",Plot), 
    ifelse(nchar(Plot) == 5, substr(Plot, 1, 2),
    ifelse(nchar(Plot) == 4, substr(Plot, 1, 1), Plot)),Block)) 


for(i in 1:12) {
# Soil properties
x[x$Block == i & x$Location == "BZ" & x$Year == "2021" & !is.na(x$Block),"pH"] = bz.soil$`1:1 Soil pH`[i]
x[x$Block == i & x$Location == "BZ" & x$Year == "2021" & !is.na(x$Block),"Nitrate"] = bz.soil$`Nitrate-N ppm N`[i]
x[x$Block == i & x$Location == "BZ" & x$Year == "2021" & !is.na(x$Block),"OM"] = bz.soil$`Organic Matter LOI %`[i]
x[x$Block == i & x$Location == "BZ" & x$Year == "2021" & !is.na(x$Block),"K"] = bz.soil$`Potassium ppm K`[i]

x[x$Block == i & x$Location == "BZ" & x$Year == "2022" & !is.na(x$Block),"pH"] = bz22.soil$`1:1 Soil pH`[i]
x[x$Block == i & x$Location == "BZ" & x$Year == "2022" & !is.na(x$Block),"Nitrate"] = bz22.soil$`Nitrate-N ppm N`[i]
x[x$Block == i & x$Location == "BZ" & x$Year == "2022" & !is.na(x$Block),"OM"] = bz22.soil$`Organic Matter LOI %`[i]
x[x$Block == i & x$Location == "BZ" & x$Year == "2022" & !is.na(x$Block),"K"] = bz22.soil$`Potassium ppm K`[i]

x[x$Block == i & x$Location == "MO" & x$Year == "2021" & !is.na(x$Block),"pH"] = mo21.soil$`1:1 Soil pH`[i]
x[x$Block == i & x$Location == "MO" & x$Year == "2021" & !is.na(x$Block),"Nitrate"] = mo21.soil$`Nitrate-N ppm N`[i]
x[x$Block == i & x$Location == "MO" & x$Year == "2021" & !is.na(x$Block),"OM"] = mo21.soil$`Organic Matter LOI %`[i]
x[x$Block == i & x$Location == "MO" & x$Year == "2021" & !is.na(x$Block),"K"] = mo21.soil$`Potassium ppm K`[i]


x[x$Block == i & x$Location == "MO" & x$Year == "2022" & !is.na(x$Block),"pH"] = mo22.soil$`1:1 Soil pH`[i]
x[x$Block == i & x$Location == "MO" & x$Year == "2022" & !is.na(x$Block),"Nitrate"] = mo22.soil$`Nitrate-N ppm N`[i]
x[x$Block == i & x$Location == "MO" & x$Year == "2022" & !is.na(x$Block),"OM"] = mo22.soil$`Organic Matter LOI %`[i]
x[x$Block == i & x$Location == "MO" & x$Year == "2022" & !is.na(x$Block),"K"] = mo22.soil$`Potassium ppm K`[i]

x[x$Block == i & x$Location == "SD" & x$Year == "2021" & !is.na(x$Block),"pH"] = sd21.soil$pH[i]
x[x$Block == i & x$Location == "SD" & x$Year == "2021" & !is.na(x$Block),"Nitrate"] = sd21.soil$`Nitrate-N`[i]
x[x$Block == i & x$Location == "SD" & x$Year == "2021" & !is.na(x$Block),"OM"] = sd21.soil$OM[i]
x[x$Block == i & x$Location == "SD" & x$Year == "2021" & !is.na(x$Block),"K"] = sd21.soil$K[i]


x[x$Block == i & x$Location == "SD" & x$Year == "2022" & !is.na(x$Block),"pH"] = sd22.soil$pH[i]
x[x$Block == i & x$Location == "SD" & x$Year == "2022" & !is.na(x$Block),"Nitrate"] = sd22.soil$`Nitrate-N`[i]
x[x$Block == i & x$Location == "SD" & x$Year == "2022" & !is.na(x$Block),"OM"] = sd22.soil$OM[i]
x[x$Block == i & x$Location == "SD" & x$Year == "2022" & !is.na(x$Block),"K"] = sd22.soil$K[i]

#tillage
x[x$Location == "BZ","tillage"] = "1"
x[x$Location != "BZ","tillage"] = "0"
x$tillage = as.factor(x$tillage)

}

#Precip
x[x$Location == "BZ" & x$Year == "2021","Precip"] = 172
x[x$Location == "BZ" & x$Year == "2022","Precip"] = 198
x[x$Location == "MO" & x$Year == "2021","Precip"] = 171
x[x$Location == "MO" & x$Year == "2022","Precip"] = 115
x[x$Location == "SD" & x$Year == "2021","Precip"] = 204
x[x$Location == "SD" & x$Year == "2022","Precip"] = 360

x$texture = ifelse(x$Location == "BZ","SiltLoam", "ClayLoam")
x$PrevCrop = ifelse(x$Location == "BZ", "fallow",ifelse(x$Location == "MO","mustard","soybean"))
x$PrevCrop2yr = x$PrevCrop
x[x$Location == "MO" & x$Year == "2021","PrevCrop2yr"] = "mustard21"
x[x$Location == "MO" & x$Year == "2022","PrevCrop2yr"] = "mustard22"

sample_data(psField) = x

sample_data(psField)$locyr = paste(sample_data(psField)$Location,substr(sample_data(psField)$Year,3,4),sep="")

saveRDS(psField,"Phyloseq2.0_Tempest/16S_maxEE_2_HI16S/ps.RDS")
```




# Step 4: Filtering
1. Load phyloseq object
2. Prevalence Filter -> Removes ASVs with only one instance across all location-years.
3. Read Depth 10k filter -> creates phyloseq object only with samples having > 10,000 reads. 
    *NAMED ps.filt.RDS*



### Prevalence > 1 THEN sample threshold of 10,000 reads
```{r Prevalence Filter THEN 10k read depth}
# Load phyloseq object
ps # 214551 taxa, 2083 samples


# Prevalence Threshold of Greater than 1
ps1 <- subset_taxa(ps, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized")) # Remove uncharacterized reads (19,194 reads total)

ps1 # 196175 taxa, 2083 samples

prevdf = apply(X = otu_table(ps1),
               MARGIN = ifelse(taxa_are_rows(ps1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})
keepTaxa = names(prevdf)[(prevdf > 1)]
ps.prev = prune_taxa(keepTaxa, ps1) 

ps.prev # 69785 taxa, 2083 samples
# Lost 126,390 taxa that only occurred once...



# READ DEPTH 10k

# Map read depth of all samples
readDepth = as.data.frame(rowSums(otu_table(ps.prev)))
readDepth$Location = as.factor(substr(row.names(readDepth),nchar(row.names(readDepth))-3,
                                      nchar(row.names(readDepth))-2))
readDepth$Year = as.factor(substr(row.names(readDepth),nchar(row.names(readDepth))-1,
                                      nchar(row.names(readDepth))))
colnames(readDepth)[1] = "Reads"

# Remove lines with < 10,000 read depth
lines10k = readDepth[readDepth$Reads >=10000,]
lines10k = merge(lines10k, sample_data(ps.prev), by = "row.names")
sample_data(ps.prev)$Row.names = row.names(sample_data(ps.prev))

ps.lines10k = ps.prev %>%
  subset_samples(Row.names %in% lines10k$Row.names)

ps.lines10k # 69785 taxa, 2024 samples
# phyloseq is dumb enough not to remove ASVs that no longer exist.

# Remove ASVs that no longer exist
dim(otu_table(ps.lines10k)[,colSums(otu_table(ps.lines10k))==0]) # 901 ASVs to remove
taxa_sums(ps.lines10k)>0
ps.filt = prune_taxa(taxa_sums(ps.lines10k)>0, ps.lines10k)

ps.filt # 69782 taxa, 2024 samples

# In total Removed:
# 144,769 taxa
# 59 samples

#saveRDS(ps.filt,"Processing/Phyloseq2.0_Tempest/16S_maxEE_2_HI16S/ps.filt.RDS")

```



# Extra: Create merged Field Data - LOAD THIS CHUNK'S OUTPUT ABOVE
```{r include=FALSE}

# Load fieldData_Combined.RDS - output of this chunk.

library(readxl)
# Field Data 2022
fieldData22 = read_excel("C:/Users/killi/OneDrive - Montana State University/Grad/Documents/Projects/2022_Field/BMB_RawFieldData.xlsx", sheet = 1)
fieldData22 = fieldData22[fieldData22$Year == "2022",]
fieldData22 = fieldData22[,c(1:21)]


# Field data 2021
fieldData21 = read_xlsx("C:/Users/killi/OneDrive - Montana State University/Grad/Documents/Projects/2021_Field/MB_2021_FieldData.xlsx", sheet = 1)
fieldData.HI = read_xlsx("C:/Users/killi/OneDrive - Montana State University/Grad/Documents/Projects/Field_Combined/hawaii_fieldData.xlsx")
fieldData21 = full_join(fieldData21,fieldData.HI)
fieldData21 = fieldData21[,1:21]

fieldData = as.data.frame(rbind(fieldData21,fieldData22))
fieldData$locyr = paste(fieldData$Location,substr(fieldData$Year,3,4),sep = "")
row.names(fieldData) = paste(fieldData$Plot,fieldData$locyr,sep = "_")


# ALIGN WITH SEQTAB
samples.out <- rownames(seqtab) # GOOD
View(as.data.frame(samples.out))
samdf <- data.frame(Subject=samples.out)

fieldData$locyrPlot = paste(fieldData$Plot,fieldData$locyr,sep="_")
fieldSampleData = merge(samdf, fieldData, by.x = "Subject", by.y = "locyrPlot", all.x = TRUE) #Merge field data with sample names to align with phyloseq object
row.names(fieldSampleData) = fieldSampleData[,1]
names(fieldSampleData)[1] = paste("Plot") 

#populate bulk samples with year, location, plot, and block information
fieldSampleData[endsWith(row.names(fieldSampleData),"22"),]$Year = "2022" 
fieldSampleData[endsWith(row.names(fieldSampleData),"21"),]$Year = "2021" 

fieldSampleData[grepl("BZ",row.names(fieldSampleData)),]$Location = "BZ"
fieldSampleData[grepl("MO",row.names(fieldSampleData)),]$Location = "MO"
fieldSampleData[grepl("SD",row.names(fieldSampleData)),]$Location = "SD"
fieldSampleData[grepl("HI",row.names(fieldSampleData)),]$Location = "HI"

fieldSampleData$Plot <- sub("_.*", "", rownames(fieldSampleData))  # extract part before "_"

fieldSampleData[grepl("B_",row.names(fieldSampleData)),]$Block = substr(row.names(fieldSampleData[grepl("B_",row.names(fieldSampleData)),]),1,nchar(row.names(fieldSampleData[grepl("B_",row.names(fieldSampleData)),]))-8) 

fieldSampleData[!grepl("B_",row.names(fieldSampleData)),]$Block = substr(row.names(fieldSampleData[!grepl("B_",row.names(fieldSampleData)),]),1,nchar(row.names(fieldSampleData[!grepl("B_",row.names(fieldSampleData)),]))-7)

fieldSampleData[grepl("B_",row.names(fieldSampleData)),]$Line = "Bulk"

#saveRDS(fieldSampleData, "C:/Users/killi/OneDrive - Montana State University/PostDoc/Microbiome_bacteria/fieldData_Combined.RDS")

```