---
title: "Processing with 16S Hawaii"
author: "Erik Killian"
date: "2025-08-05"
output: html_document
---
---
title: "phyloseq_processing"
author: "Erik Killian"
date: "2025-06-10"
output: html_document
editor_options: 
  chunk_output_type: console
---

This follows the same protocol as max_EE_2, except now I've fixed the Hawaii data that was previously a copy of the ITS data instead of 16S data.


# Step 1: Load Phyloseq Pieces
```{r LOAD HERE}
library(dada2)
library(phyloseq)
library(readxl)
library(dplyr)
library(ggplot2)

getwd()
# Loads merged field data - created in the chunk below.
fieldData = readRDS("~/phyloseq/agData/fieldData_Combined.RDS")

# Loaded from tempest output 
## 2021.R and 2022.R create seqtabs 
## merging.R merge seqtabs, remove chimeras, and creates taxtab
## Phylogenetic_Tree.R uses merging.R output "seqtab_NoC" to create phylo tree

## seqtab is the merged seqtabs from each location-year, merged then chimeras removed.
## taxtab is the taxonomy table, made from seqtab to classify taxonomy.

seqtab = readRDS("~/phyloseq/16S/seqtab.merged.NoC.RDS")
taxtab = readRDS("~/phyloseq/16S/taxTab.RDS")

```

# Step 2: Merge Phyloseq into One

```{r Create phyloseq object}
ps = phyloseq(otu_table(seqtab,taxa_are_rows = FALSE),
              tax_table(taxtab),
              sample_data(fieldData))


# Add ASV short names
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps <- merge_phyloseq(ps, dna)
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
ps.all = ps
ps = subset_taxa(ps, Kingdom == "Bacteria") # Removes only 403 taxa 

saveRDS(ps,"ps.RDS")

table(taxtab[,"Kingdom"])
table(tax_table(ps)[,"Kingdom"])
table(sample_data(ps)[,"Location"])

```


# Step 3: Filtering
1. Load phyloseq object
2. Prevalence Filter -> Removes ASVs with only one instance across all location-years.
3. Read Depth 10k filter -> creates phyloseq object only with samples having > 10,000 reads. 
    *NAMED ps.filt.RDS*



### Prevalence > 1 THEN sample threshold of 10,000 reads
```{r Prevalence Filter THEN 10k read depth}
# Load phyloseq object
ps # 214551 taxa, 2083 samples


# Prevalence Threshold of Greater than 1
ps1 <- subset_taxa(ps, !is.na(Phylum) & !Phylum %in% c("", "uncharacterized")) # Remove uncharacterized reads (19,194 reads total)

ps1 # 196175 taxa, 2083 samples

prevdf = apply(X = otu_table(ps1),
               MARGIN = ifelse(taxa_are_rows(ps1), yes = 1, no = 2),
               FUN = function(x){sum(x > 0)})
keepTaxa = names(prevdf)[(prevdf > 1)]
ps.prev = prune_taxa(keepTaxa, ps1) 

ps.prev # 69785 taxa, 2083 samples
# Lost 126,390 taxa that only occurred once...



# READ DEPTH 10k

# Map read depth of all samples
readDepth = as.data.frame(rowSums(otu_table(ps.prev)))
readDepth$Location = as.factor(substr(row.names(readDepth),nchar(row.names(readDepth))-3,
                                      nchar(row.names(readDepth))-2))
readDepth$Year = as.factor(substr(row.names(readDepth),nchar(row.names(readDepth))-1,
                                      nchar(row.names(readDepth))))
colnames(readDepth)[1] = "Reads"

# Remove lines with < 10,000 read depth
lines10k = readDepth[readDepth$Reads >=10000,]
lines10k = merge(lines10k, sample_data(ps.prev), by = "row.names")
sample_data(ps.prev)$Row.names = row.names(sample_data(ps.prev))

ps.lines10k = ps.prev %>%
  subset_samples(Row.names %in% lines10k$Row.names)

ps.lines10k # 69785 taxa, 2024 samples
# phyloseq is dumb enough not to remove ASVs that no longer exist.

# Remove ASVs that no longer exist
dim(otu_table(ps.lines10k)[,colSums(otu_table(ps.lines10k))==0]) # 901 ASVs to remove
taxa_sums(ps.lines10k)>0
ps.filt = prune_taxa(taxa_sums(ps.lines10k)>0, ps.lines10k)

ps.filt # 69782 taxa, 2024 samples

# In total Removed:
# 144,769 taxa
# 59 samples

#saveRDS(ps.filt,"ps.filt.RDS")



```
